Unsupervised acquisition of comprehensive multiword lexicons \\ using competition in an $n$-gram lattice

### 2016-09-28 eacl2017.tex
Formatting tweaks

\documentclass[11pt]{article}
\usepackage{eacl2017}
\usepackage{times}
\usepackage{url}
\usepackage{latexsym}
\usepackage[pdftex]{graphicx}

@@

@@

In this paper we are attempting to identify \emph{formulaic sequences}, following the terminology of Wray \shortcite{Wray02,Wray08}, who defines a formulaic sequence (FS) as ``a sequence, continuous or discontinuous, of words or other elements, which is, or appears to be, prefabricated: that is, stored and retrieved whole from memory at the time of use, rather than being subject to generation or analysis by the language grammar.'' In other words, a formulaic sequence shows signs of being lexicalized. In computational linguistics, the most common term used to describe multiword lexical units is \emph{multiword expression} or MWE \cite{Baldwin10}, but here we wish to make a principled distinction between at least somewhat non-compositional, strongly lexicalized MWEs and FS, a superset which includes MWEs but also (apparently) compositional linguistic formulas. This distinction is not a new one, for instance it exists to some extent in the MWE annotation of Schneider \cite{Schneider14a}, who distinguishes between weak (collocational) and strong (non-compositional) MWEs, but it is our contention is that separate, precise terminology is useful for research targeted at either class: we need not strain the concept of MWE to include items which do not require special semantics, nor are we included to disregard or minimize the larger formulaticity of language simply because it is not the dominant focus of MWE research. Many MWE researchers would defensibly balk at including in their MWE lexicons and corpus annotations (English) formulaic sequences such as \emph{there is something going on}, \emph{it is more important than ever to\ldots}, \emph{\ldotsdo not know what it is like to\ldots},\emph{it would be great if * could}, \emph{there is no shortage of\ldots}, \emph{the rise and fall of\ldots}, \emph{now is not the time to\ldots}, \etc as well as tens of thousands of other such phrases which, along with less compositional, clearly MWE formulas like \emph{be worth * weight in gold}, are identified by the FS-extraction system presented here.

\begin{figure}[!tb]
\center{\includegraphics{FS_diagram_1.png}}
\caption{Multiword Terminology}

@@

Our approach to FS identification involves optimization of the total cost of a lattice where each node corresponds to an $n$-gram type derived from a large corpus. The cost of the whole lattice is defined simply as a product of the costs of the individual nodes. Each node can be considered either ``on'' (is an FS) or ``off'' (is not an FS), with the base cost of being off simply a variation on the LPR association measure. Most fundamentally, $n$-gram nodes in the lattice are directionally connected to nodes consisting of $n+1$-grams which subsume them and $n-1$-grams which they subsume: that is, the (gapped) $n$-gram \emph{keep * under wraps} would, for instance be connected ``upwards'' to the node \emph{keep everything under wraps} and connected ``downwards'' to \emph{under wraps}. These directional relationships allow for two basic effects between nodes in the lattice when a node is turned on: \emph{covering}, which inhibits nodes below (subsumed by) a turned-on node (\eg, if \emph{keep * under wraps} is on, the model will tend not to choose \emph{under wraps} as an FS); and \emph{clearing}, which inhibits nodes above a turned-on node (\eg, if \emph{keep * under wraps} is on, the model would avoid selecting \emph{keep everything under wraps} as an FS). A third, undirected mechanism is \emph{overlapping}, where nodes inhibit each other due to (non-subsuming) overlaps in the corpus (\eg having both \emph{keep * under wraps} and \emph{be keep * under} as FS will be avoided). Given these relationships, we minimize the cost of the entire lattice by iterating over the nodes and greedily optimizing the on/off choice relative to the local neighborhood of each node, until convergence.

@@

Once a set of relevant $n$-grams is identified (and counted), other statistics required to calculate \emph{Lexical Predictability Ratios} (LPR) for each word in the $n$-gram are collected. Again, LPR measures how statistically unlikely a particular word is relative to how predictable the word is given the same amount of syntactic context. Formally, the LPR for some word $w_i$ in the context of a sequence of words of length $n$ which have a corresponding set of sequence of POS tags $t$, is given by:

@@

The use of minimum here should be understood in the context of competition between nodes in the lattice: minLPR for a particular $n$-gram does not reflect the \emph{overall}
degree to which an $n$-gram holds together, instead it focuses on the word which is its weakest link. For example, in the case we are evaluating the quality of \emph{be keep * under} as an FS, a general statistical metric might give it a high score due the strong statistical relationship between \emph{keep} and \emph{under}, but the minLPR is focused on the weaker relationship between \emph{be} and \emph{keep * under}.

@@

The most important node interaction is \textit{covering}, which corresponds to discounting or entirely excluding a node based on the presence of a node higher in the lattice. Our model includes two types of covering: hard and soft. Hard covering is based on the idea that, due to very similar counts, we can reasonably conclude that the presence of $n$-gram in our statistics is a direct result of the other: if we have 143 counts of \emph{keep * under wraps} and 152 counts of \emph{under wraps}, the presence of \emph{keep * under wraps} almost completely explains \emph{under wraps}, and it is better if we collapse these two decisions into one. We do this by permanently disabling any covered node, and giving the covering node the maximum minLPR among all the nodes it covers; this means that longer $n$-grams with function words (which often have lower minLPR) can benefit from the strong statistical relationships between open-class lexical features in $n$-grams that they cover. This is done as a preprocessing step, and besides being effective it also greatly improves the tractability of the iterative optimization of the lattice. Of course, a threshold for hard covering must be chosen: empirically, we have found that a ratio of $2/3$ (corresponding to a significant but not overwhelming majority of the counts of a lower node corresponding to the higher node) works well. Finally, we also use hard covering to partially address the rather challenging problem of pronouns, which, due to pragmatic biases in particular corpora, often have higher than desirable LPR ratios, but cannot reasonably be excluded either since many common FS contain them. In our model, $n$-grams with pronouns are considered covered (inactive) unless they cover at least one other node which does not have a pronoun.

Soft covering is used in cases when a single $n$-gram does not entirely account for another, but a turned-on $n$-gram to some extent may explain some of the statistical irregularity of one lower in the lattice. For instance, \emph{keep * under} is not hard-covered by \emph{keep * under wraps} (since there are FS such as \emph{keep * under surveillance}, \emph{keep it under your hat}, \etc), but if \emph{keep * under wraps} is tagged as an FS, we nevertheless want to discount the portion of the \emph{keep * under} counts that correspond to \emph{keep * under wraps}. We accomplish this by lowering the turned-off cost function of \emph{keep * under} (and thus making turning on less desirable) in the following manner: let $t$ be the target node, and $s_1\ldots_m$ be any turned-on nodes which are above $t$ in the lattice. Then, then the new $d_0$, $d_0'$ for a node $t$ is:

\begin{displaymath}
d_{0}'(t) = \frac{c(t) - \sum{c(s_1)\ldots c(s_m)}}{c(t)} d_0(t)

\end{displaymath}

@@

In general, covering prefers turning on longer, covering $n$-grams since doing so lowers the turned-off cost function of nodes lower in the lattice. Not surprisingly, it is important to have a mechanism working in opposition, \ie one which generally prefers shorter $n$-grams. \textit{Clearing} does this by changing the $C_0$ of nodes higher in the lattice when a lower node is turned-on. The basic mechanism is the same as covering, except that we cannot make direct make use of counts in the same way; whereas it makes sense to discount the turned-off cost of covered nodes in proportion to the (lower) relative counts of their covering nodes, in the reverse direction this logic entirely fails. A simple but effective solution is make use of the minLPR value: in this paper, we consider two options: a hard clearing where any turned-on node will clear (set $C_{0}$ to 0 for) any node higher in the lattice which has a minLPR that is lower than it; and a soft clearing which lowers  $C_{0}(t)$ based on the ratio between the minLPR for the two relevant nodes, \ie

\begin{displaymath}
C_{0}'(t) = max(1, C_{0}(t) \prod{\frac{minLPR(t)}{minLPR(s_i)})}

\end{displaymath}
\noindent
where $s_i$ is the $i$th clearing node. We refer to this mechanism as clearing because it tends to clear away a variety of trivial uses of common FS which many have higher LPR due to the lexical and syntactic specificity of the FS. To give an example of its effect, if the node \emph{keep * under wraps} is turned on, then, assuming that their minLPR is lower, under hard clearing the $C_0$ of nodes such as \emph{be keep * under wraps} and \emph{keep * under wraps for} will be set to 1. This dissuades (but does not prevent) the model from turning on these nodes. Note that the effect of both covering and clearing is to lower the cost function for covered/clears nodes, which encourages the model to turn an $n$-gram like \emph{keep * under wraps} on, even supposing its minLPR were relatively low.

The third mechanism of node interaction involves $n$-grams which overlap in the corpus. Though there are some token-level exceptions (\eg coordinated compound nouns), it is generally the case that independent formulaic sequences do not overlap. For example, given that \emph{be keep * under} and \emph{keep * under wraps} often appear together, overlapping on the tokens \emph{keep * under}, we do not want both being selected as an FS, even in the case where both had high minLPR. To address this problem, we use a mechanism somewhat different than above: rather than lowering the off-cost, we raise the on-cost. Let $oc(x,y)$ refer to number of times nodes $x$ and $y$ overlapped in the corpus, and $o_1\ldots o_m$ refer to \emph{turned-on} nodes which overlap with our target node $t$. With analogy to $d_0$ above, we define an $d_1$ exponential term for our target node $t$ as:

@@

Given the relationships between nodes that we have defined above, it is obviously not sufficient to consider switching only the present node; If, for instance, one or more of \emph{be keep * under wraps}, \emph{under wraps}, or \emph{be keep * under} has been turned on, the covering, blocking, or overlapping effects of these other nodes will likely prevent a competing node like \emph{keep * under wraps} from being correctly activated. Instead, the algorithm identifies a small set of nodes which are the most relevant to status of the node under consideration whose status should be considered simultaneously. Since turned-off nodes have no direct effect on each other, only turned-on nodes above, below, or overlapping with the current node need be considered.  Once the relevant nodes have been identified, all (including turned-off nodes) nodes whose cost function is affected by one or more of the relevant nodes are identified, and then a greedy search is carried out for the optimal configuration of the relevant nodes starting from an `all-on' state.

@@

>>>>

\documentclass[11pt]{article}
\usepackage{eacl2017}
\usepackage{times}
\usepackage{url}
\usepackage{xspace}
\usepackage{latexsym}
\usepackage[pdftex]{graphicx}

@@

% TJB macros
\newcommand{\dotts}{...}
\newcommand{\gap}{$*$\xspace}
\newcommand{\ex}[1]{\textit{#1}\xspace}
\newcommand{\termdef}[1]{\textbf{#1}\xspace}
\newcommand{\termemph}[1]{\textit{#1}\xspace}

@@

In this paper we are attempting to identify \termdef{formulaic sequences}, following the terminology of Wray \shortcite{Wray02,Wray08}, who defines a formulaic sequence (FS) as ``a sequence, continuous or discontinuous, of words or other elements, which is, or appears to be, prefabricated: that is, stored and retrieved whole from memory at the time of use, rather than being subject to generation or analysis by the language grammar.'' In other words, a formulaic sequence shows signs of being lexicalized. In computational linguistics, the most common term used to describe multiword lexical units is \emph{multiword expression} or MWE \cite{Baldwin10}, but here we wish to make a principled distinction between at least somewhat non-compositional, strongly lexicalized MWEs and FS, a superset which includes MWEs but also (apparently) compositional linguistic formulas. This distinction is not a new one, for instance it exists to some extent in the MWE annotation of Schneider \cite{Schneider14a}, who distinguishes between weak (collocational) and strong (non-compositional) MWEs, but it is our contention is that separate, precise terminology is useful for research targeted at either class: we need not strain the concept of MWE to include items which do not require special semantics, nor are we included to disregard or minimize the larger formulaticity of language simply because it is not the dominant focus of MWE research. Many MWE researchers would defensibly balk at including in their MWE lexicons and corpus annotations (English) formulaic sequences such as \ex{there is something going on}, \ex{it is more important than ever to \dotts}, \ex{\dotts do not know what it is like to \dotts}, \ex{it would be great if \gap could}, \ex{there is no shortage of\dotts}, \ex{the rise and fall of\dotts}, \ex{now is not the time to\dotts}, \etc as well as tens of thousands of other such phrases which, along with less compositional, clearly MWE formulas like \ex{be worth \gap weight in gold}, are identified by the FS-extraction system presented here.

\begin{figure}[!t]
\center{\includegraphics[width=0.35\textwidth]{FS_diagram_1.pdf}}
\caption{Multiword Terminology}

@@

Our approach to FS identification involves optimization of the total cost of a lattice where each node corresponds to an $n$-gram type derived from a large corpus. The cost of the whole lattice is defined simply as a product of the costs of the individual nodes. Each node can be considered either ``on'' (is an FS) or ``off'' (is not an FS), with the base cost of being off simply a variation on the LPR association measure. Most fundamentally, $n$-gram nodes in the lattice are directionally connected to nodes consisting of $n+1$-grams which subsume them and $n-1$-grams which they subsume: that is, the (gapped) $n$-gram \ex{keep \gap under wraps} would, for instance be connected ``upwards'' to the node \ex{keep everything under wraps} and connected ``downwards'' to \ex{under wraps}. These directional relationships allow for two basic effects between nodes in the lattice when a node is turned on: \termdef{covering}, which inhibits nodes below (subsumed by) a turned-on node (\eg, if \ex{keep \gap under wraps} is on, the model will tend not to choose \ex{under wraps} as an FS); and \termdef{clearing}, which inhibits nodes above a turned-on node (\eg, if \ex{keep \gap under wraps} is on, the model would avoid selecting \ex{keep everything under wraps} as an FS). A third, undirected mechanism is \termdef{overlapping}, where nodes inhibit each other due to (non-subsuming) overlaps in the corpus (\eg having both \ex{keep \gap under wraps} and \ex{be keep \gap under} as FS will be avoided). Given these relationships, we minimize the cost of the entire lattice by iterating over the nodes and greedily optimizing the on/off choice relative to the local neighborhood of each node, until convergence.

@@

Once a set of relevant $n$-grams is identified (and counted), other statistics required to calculate \termdef{Lexical Predictability Ratios} (LPR) for each word in the $n$-gram are collected. Again, LPR measures how statistically unlikely a particular word is relative to how predictable the word is given the same amount of syntactic context. Formally, the LPR for some word $w_i$ in the context of a sequence of words of length $n$ which have a corresponding set of sequence of POS tags $t$, is given by:

@@

The use of minimum here should be understood in the context of competition between nodes in the lattice: minLPR for a particular $n$-gram does not reflect the \emph{overall}
degree to which an $n$-gram holds together, instead it focuses on the word which is its weakest link. For example, in the case we are evaluating the quality of \ex{be keep \gap under} as an FS, a general statistical metric might give it a high score due the strong statistical relationship between \ex{keep} and \ex{under}, but the minLPR is focused on the weaker relationship between \ex{be} and \ex{keep \gap under}.

@@

The most important node interaction is \termdef{covering}, which corresponds to discounting or entirely excluding a node based on the presence of a node higher in the lattice. Our model includes two types of covering: hard and soft. Hard covering is based on the idea that, due to very similar counts, we can reasonably conclude that the presence of $n$-gram in our statistics is a direct result of the other: if we have 143 counts of \ex{keep \gap under wraps} and 152 counts of \ex{under wraps}, the presence of \ex{keep \gap under wraps} almost completely explains \ex{under wraps}, and it is better if we collapse these two decisions into one. We do this by permanently disabling any covered node, and giving the covering node the maximum minLPR among all the nodes it covers; this means that longer $n$-grams with function words (which often have lower minLPR) can benefit from the strong statistical relationships between open-class lexical features in $n$-grams that they cover. This is done as a preprocessing step, and besides being effective it also greatly improves the tractability of the iterative optimization of the lattice. Of course, a threshold for hard covering must be chosen: empirically, we have found that a ratio of $2/3$ (corresponding to a significant but not overwhelming majority of the counts of a lower node corresponding to the higher node) works well. Finally, we also use hard covering to partially address the rather challenging problem of pronouns, which, due to pragmatic biases in particular corpora, often have higher than desirable LPR ratios, but cannot reasonably be excluded either since many common FS contain them. In our model, $n$-grams with pronouns are considered covered (inactive) unless they cover at least one other node which does not have a pronoun.

Soft covering is used in cases when a single $n$-gram does not entirely account for another, but a turned-on $n$-gram to some extent may explain some of the statistical irregularity of one lower in the lattice. For instance, \ex{keep \gap under} is not hard-covered by \ex{keep \gap under wraps} (since there are FS such as \ex{keep \gap under surveillance}, \ex{keep it under your hat}, \etc), but if \ex{keep \gap under wraps} is tagged as an FS, we nevertheless want to discount the portion of the \ex{keep \gap under} counts that correspond to \ex{keep \gap under wraps}. We accomplish this by lowering the turned-off cost function of \ex{keep \gap under} (and thus making turning on less desirable) in the following manner: let $t$ be the target node, and $s_1\dotts_m$ be any turned-on nodes which are above $t$ in the lattice. Then, then the new $d_0$, $d_0'$ for a node $t$ is:

\begin{displaymath}
d_{0}'(t) = \frac{c(t) - \sum{c(s_1)\ldots c(s_m)}}{c(t)} d_0(t)
\end{displaymath}

@@

In general, covering prefers turning on longer, covering $n$-grams since doing so lowers the turned-off cost function of nodes lower in the lattice. Not surprisingly, it is important to have a mechanism working in opposition, \ie one which generally prefers shorter $n$-grams. \termdef{Clearing} does this by changing the $C_0$ of nodes higher in the lattice when a lower node is turned-on. The basic mechanism is the same as covering, except that we cannot make direct make use of counts in the same way; whereas it makes sense to discount the turned-off cost of covered nodes in proportion to the (lower) relative counts of their covering nodes, in the reverse direction this logic entirely fails. A simple but effective solution is make use of the minLPR value: in this paper, we consider two options: a hard clearing where any turned-on node will clear (set $C_{0}$ to 0 for) any node higher in the lattice which has a minLPR that is lower than it; and a soft clearing which lowers  $C_{0}(t)$ based on the ratio between the minLPR for the two relevant nodes, \ie

\begin{displaymath}
C_{0}'(t) = max(1, C_{0}(t) \prod{\frac{minLPR(t)}{minLPR(s_i)})}
\end{displaymath}
where $s_i$ is the $i$th clearing node. We refer to this mechanism as clearing because it tends to clear away a variety of trivial uses of common FS which many have higher LPR due to the lexical and syntactic specificity of the FS. To give an example of its effect, if the node \ex{keep \gap under wraps} is turned on, then, assuming that their minLPR is lower, under hard clearing the $C_0$ of nodes such as \ex{be keep \gap under wraps} and \ex{keep \gap under wraps for} will be set to 1. This dissuades (but does not prevent) the model from turning on these nodes. Note that the effect of both covering and clearing is to lower the cost function for covered/clears nodes, which encourages the model to turn an $n$-gram like \ex{keep \gap under wraps} on, even supposing its minLPR were relatively low.

The third mechanism of node interaction involves $n$-grams which overlap in the corpus. Though there are some token-level exceptions (\eg coordinated compound nouns), it is generally the case that independent formulaic sequences do not overlap. For example, given that \ex{be keep \gap under} and \ex{keep \gap under wraps} often appear together, overlapping on the tokens \ex{keep \gap under}, we do not want both being selected as an FS, even in the case where both had high minLPR. To address this problem, we use a mechanism somewhat different than above: rather than lowering the off-cost, we raise the on-cost. Let $oc(x,y)$ refer to number of times nodes $x$ and $y$ overlapped in the corpus, and $o_1\dotts o_m$ refer to \termemph{turned-on} nodes which overlap with our target node $t$. With analogy to $d_0$ above, we define an $d_1$ exponential term for our target node $t$ as:

@@

Given the relationships between nodes that we have defined above, it is obviously not sufficient to consider switching only the present node; If, for instance, one or more of \ex{be keep \gap under wraps}, \ex{under wraps}, or \ex{be keep \gap under} has been turned on, the covering, blocking, or overlapping effects of these other nodes will likely prevent a competing node like \ex{keep \gap under wraps} from being correctly activated. Instead, the algorithm identifies a small set of nodes which are the most relevant to status of the node under consideration whose status should be considered simultaneously. Since turned-off nodes have no direct effect on each other, only turned-on nodes above, below, or overlapping with the current node need be considered.  Once the relevant nodes have been identified, all (including turned-off nodes) nodes whose cost function is affected by one or more of the relevant nodes are identified, and then a greedy search is carried out for the optimal configuration of the relevant nodes starting from an `all-on' state.

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
